---
title: "GOA northen rockfish FT-NIRS"
author: Ben Williams
date: today
format: 
  html:
    code-fold: true
    self-contained: true
execute:
  message: false
  warning: false
    
---

All code and data can be found on [github](https://github.com/BenWilliams-NOAA/ftnirs). 

```{r}
# load ----
library(tidytable)
library(ggplot2)
library(here)
library(vroom)
library(readxl)
library(PBSmapping)
library(flextable)
```

## Initial data examination

To start I downloaded data from the NOAA fisheries [In Port](https://www.fisheries.noaa.gov/inport/item/73040).
These data are in the `data` folder of this repo.

```{r}
vroom(here::here("data","AGP_MMCNN_NRF2013to2019.csv")) %>% 
  select(sample, final_age, latitude, longitude, length, sex) -> agp 
glimpse(agp) 
```

However, more associated data are needed to generate age compositions for the GOA northern rockfish assessment. 
I received the same data with associated survey cruise numbers and haul id from Jon Short 2024-12-18.

```{r}
read_excel(here::here("data", "nrf_traintunetest_predictions_with_haul.xlsx")) %>%
            rename(vessel=vessel_code, specimenid=specimen_number) -> dat # update has haul id
glimpse(dat)
```

To align all of the necessary data columns I pulled in all survey locations from the GOA that include northern rockfish (code for this is in the `data_call.R` script), then filter the FT-NIRS data to only include GOA samples as there are AI samples as well.

```{r}
vroom(here::here("data", "goa_specimen_data.csv")) -> bts # GOA bottom trawl survey all sites
dat %>% 
  left_join(bts) %>% 
  filter(!is.na(start_longitude)) -> dat
# glimpse(dat)
```

```{r}
#| fig.cap: "Red points are locations for Aleutian Island samples, blue point are from the Gulf of Alaska."
# map data
data('nepacLLhigh')
nepacLLhigh %>%
  dplyr::select(group=PID, POS=POS,long=X, lat=Y) -> ak

ggplot() +
  geom_polygon(data = ak, aes(long, lat, group = group), fill=8, color='black') +
  theme(panel.background = element_rect(fill = 'white')) +
  xlab(expression(paste(Longitude^o,~'W'))) +
  ylab(expression(paste(Latitude^o,~'W'))) +
  geom_point(data = agp, aes(longitude, latitude), color = 'red') +
  geom_point(data = bts, aes(start_longitude, start_latitude), color = 'blue', alpha = 1) +
  coord_map(xlim = c(-180, -145), ylim = c(50, 61))
```

Removing Aleutian Island samples results in a sample size of 1,553.

```{r}
#| tbl-cap: "Number of ages (n) and number of hauls by year available from FT-NIRS analysis for the GOA northern rockfish assessment."
dat %>% 
  distinct(haul, year) %>% 
  summarise(hauls = n(), .by = year) -> hls

  dat %>% 
  summarise(n = n(), .by = year) %>% 
    left_join(hls) %>% 
  bind_rows(summarise(., across(c(n, hauls), sum), across(year, ~"Total"))) %>% 
  flextable()
```

For comparison the last full assessment (2024) had more years and hauls. 

```{r}
#| tbl-cap: "Number of ages (n) and number of hauls by year used in the 2024 GOA northern rockfish assessment."
data.frame(year = as.character(c(1990, 1993, 1996, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 
                    2017, 2019, 2021, 2023)),
           n = c(331, 242, 462, 278, 466, 216, 417, 605, 651, 430, 495, 465, 462, 368, 512, 572),
           hauls = c(12, 17, 19, 27, 85, 22, 72, 82, 69, 74, 68, 56, 80, 64, 68, 48)) %>% 
  bind_rows(summarise(., across(c(n, hauls), sum), across(year, ~"Total"))) %>% 
  flextable()

```

 - In the years that have been evaluated with FT-NIRS there are similar numbers of hauls and samples, however more samples from additional years will be necessary to fully examine incorporating these data into the assessment. 
 
## Next steps

 - Establish aging error matrix
   + Is there an age-error matrix currently available, if not can we set one up?
   + The methods in [Hulson and Williams 2024](https://www.sciencedirect.com/science/article/pii/S0165783623002874?via%3Dihub) are probably not directly applicable, as we re-sampled from multiple age reads from different people, my (limited) understanding of FT-NIRS is that this is not how aging error would be determined? An adjustment may be needed to accurately account for this error.
  - Initiate input sample size for samples in hand (bootstrap re-sampling based upon haul and sex, based on Hulson and Williams 2024). Now that I have the bottom trawl data linked with the FT-NIRS ages I can get started on this. 
  - Could more otoliths be aged for additional years? The higher the sample size the better, with a particular focus on keeping the number of hauls sampled from high.

